\TITLE

### 2020-12-19 response-v3-content.tex

\newcommand{\DONE}{{\color{green!60!black}\makebox[0pt][l]{$\square$}\raisebox{.15ex}{\hspace{0.1em}$\checkmark$}}~}
\newcommand{\DOIT}{{\color{red!60!black}\makebox[0pt][l]{$\square$}\raisebox{.15ex}{\hspace{0.1em}$\boxtimes$}}~}
\newcommand{\PROGRESS}{(in progress)~}

@@

REVISION DUE: 01.12.2021

\section{Additional actions taken}

    \item \DONE We used the new JDS \LaTeX template available from the Journal page instead of the older one that we used.

    \item \DONE Due to that all references are now in the style of the journal.

    \item \DONE URLs have been added with the standard urlbst tool so they can be used in the refernces. The jdsurl.bst is included in the source.

### 2020-12-19 vonLaszewski-covid.tex
add urls to bst
inline todonotes

\usepackage{calc}
\usepackage{comment}
\usepackage{todonotes}

@@

In this study, we used a Long Short-Term Memory (LSTM) algorithm for
our predictions \cite{www-keras-lstm,Hochreiter1997-dk}. An LSTM is a
recurrent neural network (RNN) \cite{Rumelhart1986-li} with feedback
connections allowing the use of data input sequences to predict data
output sequences. LSTMs have been applied to many application areas
from handwriting \cite{Graves2009-qb}, image, feature detection, and
time series prediction \cite{Schmidhuber2005-oy}. LSTM's have an
internal state that is used to prevent the vanishing gradient problem
\cite{Hochreiter1991-mp} during the training of RNNs.

@@

We often think of the laws of physics described by operators that
evolve the system given sufficient initial conditions and in this
language, we have shown how to represent Newton’s law operator by a
recurrent network \cite{Kadupitiya2020-zq}. We expect that the time
dependence of many complex systems: Covid pandemics, Southern
California earthquakes, traffic flow, security events can be described
by deep learning operators that both capture the dynamics and allow
predictions.

@@

\begin{requirement}{\bf Compute Resource Mapping.}
  For other uses of AICov, we need to integrate a flexible resource
  utilization framework. \Solution To address this requirement, we
  leverage from our earlier work and utilize through Cloudmesh and
  NIST NBDIF definitions cloud services into the architecture
  \cite{las-19-nist}. This includes containers but also infrastructure
  services in the cloud. Through these service integrations, we also
  provide access to GPUs.
\end{requirement}

@@

In summary, the AICov architecture allows the inclusion of new data
sources and models. An orchestrator enables modifying parameters and
selecting data sources and models used in the analysis. The framework
can be accessed via the interface layer that includes APIs, REST, and
Jupyter Notebooks. Compute interfaces are provided either by direct
access of GPUs from local resources or through the staging of
automatically generated REST services via our Cloudmesh REST Service
integrator
\cite{cloudmesh-openapi-install,cloudmesh-openapi-benchmark,cloudmesh-manual}. Additionally,
we can also access through abstractions cloud services that are
offered by the various cloud providers and are targeting our
application domain. Different models can be easily integrated into our
framework to verify our forecast or enhance our forecast based on the
users' models and data.

@@

While the U.S. has seen an overall case fatality rate of 5.14\%, some
U.S. locations have experienced a disproportionate number of deaths
compared to others. This disparity of deaths among different
communities could be attributed to a diversity of local risk factors
and socioeconomic determinants of health that now constitute a very
active area of investigation. Researchers are studying the effects of
different community-specific socioeconomic variables, underlying
health conditions and comorbidities, infrastructure, and systemic
responses.  Given the complexity of such underlying conditions that
can determine individual disease outcomes, where and under what
conditions people get infected by the virus, and then recover or die
take place not in isolation but rather in an interplay of diverse
factors that are characteristic to a local community, which need to be
synthesized. In other words, a basic requirement for effective
modeling of COVID-19 dynamics is that the information regarding a
particular community’s vulnerabilities and strengths pertaining to the
disease must be as comprehensive and integrative as possible.  The
regular reporting of accurate Coronavirus outbreaks data from local
health departments has been understandably difficult, especially in
the areas that are hardest hit. We use data on community-specific
health and behavioral risk factors aggregated at the county level from
the large-scale longitudinal CDC BRFSS surveys \cite{www-cdc-brfss},
and demographic and socioeconomic data from the U.S. Census Bureau
that are known at the local levels along with the cases and deaths
data compiled by NCHS \cite{www-cdc-nchs} and integrate them with
other data sources.

@@

Among the different sources of data on daily cases and deaths data
collected and aggregated at different spatial levels, especially for
the U.S., many are either freely available repositories such as GitHub
or accessible via specifically developed online resources such as the
New York Times~\cite{github-nytimes} and the Johns Hopkins Coronavirus
Resource Center~\cite{www-jh-covid19}.  For this study, we generated
two data sets based on the latter resource upon our checking for the
locations and times at which cumulative counts of COVID-19 cases and
deaths on any given day were no less than those on the previous
day. This led us to compile daily coronavirus cases and deaths data
for 345 U.S. counties from 32 states and the District of Columbia and
matched by five-digit FIPS code or county name to dynamic and static
variables from additional data sources.

The static data in this repository was collected from the U.S. Census
Bureau and the Centers for Disease Control and Prevention (CDC). The
different fields cover the domains of behavioral and health risk
factors, hospital capacity, and socioeconomic and demographic
conditions. Data on health outcomes, prevention measures, and
unhealthy behaviors were extracted from the CDC's 500 Cities Local
Data for Better Health program \cite{www-cdc-chronic-data} in the form
of crude or age-adjusted prevalence of conditions such as respiratory
disease, obesity or smoking. These values were aggregated from the
census tract to the county level using the first five digits of the
eleven-digit tract FIPS code. Because a single county may consist of
multiple individual cities, we include the list of all city labels
within each aggregate group to represent a greater metropolitan
area. 110 of such metropolitan areas that had more than 500 reported
cases of COVID-19 by April 15, 2020, were selected for this study.

@@

Demographic variables were gathered from the Census
QuickFacts~\cite{www-census} online resource using an automated web
scraping algorithm and cover relevant areas such as age, race, income,
and population density. Additional socioeconomic variables include the
Gini Index, which measures economic inequality and CDC Social
Vulnerability Index (SVI)~\cite{www-cdc-svi}. The SVI was created to
guide public health officials and disaster response efforts by
identifying the communities across the United States most likely to
need support during a crisis. Census tracts and corresponding counties
are ranked across 15 social factors, which are grouped into four
themes: socioeconomic status, household composition and disability,
minority status and language, and housing and transportation.

Lastly, the total number of general acute care, critical access, and
military hospitals within each county are included in the data. Such
variables included the number of relevant hospitals per county, and
the estimated number of beds (total known bed counts added to the
number of hospitals in the county with missing data times the average
number of beds per hospital in that state) per 1,000 people using the
American Hospital Directory~\cite{www-ahd}.

@@

To establish the feasibility of using deep learning, we have, in our
analysis, verified that through deep learning we can recreate the
predictions identified by a model-based approach using empirical
fits. In this comprehensive analysis, we obtain predictions through
deep learning while using data generated from epidemiological models
as discussed in \cite{marsland20-covid-paper}. The time series we used
was 100 days long, and a multi-layered Long Short-Term Memory (LSTM)
recurrent network was used. Our prediction approach differs by
learning not only from the demographics (fixed data for each city) and
time-dependent data but by integrating the population model for the
underlying prediction, as shown in Figure \ref{fig:magic-1}. Such model
predictive integration capability is important in any application with
multiple time scales. For example, this allows us to integrate
multi-scale time effects into the forecast, which could address the
combination of general forecasts and the next time steps of choice,
such as days, weeks, months, or longer. For this analysis, we used 37
of the 110 cities with reliable empirical fits (that we derived from
an epidemiological model and not deep learning
\cite{marsland20-covid-paper}) for the cases and deaths data up to
April 15, 2020. Our analysis identified a sophisticated single deep
learning time evolution operator that can describe these 37 separate
data sets, and smooth fitted data leads to very accurate deep learning
descriptions via LSTM as depicted in both Figure \ref{fig:magic-1}
that are smaller than 1\% from the model predictions.

@@

To answer these questions, we chose for our next analysis a date range
between 2020-02-01 and 2020-05-25. We start with a basic LSTM operator
with two layers of LSTM, and initial and final fully connected layers
\cite{Kadupitiya2020-zq}.  For LSTM, we have selected the activation
function based on Rectified Linear Units (RELU). For the last layer
sigmoid is chosen. A drop out rate of 0.2 is used, we use 16 input
nodes, and have a batch size of 110. The maximum epoch is chosen to be
200. The number of nodes internally is 32. The maximum total number of
Samples is 12540 when using 3 days as input length. We use a two-layer
LSTM, as shown in Table \ref{tab:model}. Within our system, we have
the ability to set a number of input days. For this analysis, we have
chosen three different input lengths, namely 3, 4, and 5 days.

@@

While this study focuses on if and which risk factors can improve the
predictions, we do not focus on why. Such analysis could provide
insights into causal relationships. For example, information such as
health insurance and the number of hospitals and beds are often
related to the quality of services that affect early detection and
treatment. Furthermore, population parameters such as density can
determine the rate of spread and the number of cases. Similarly,
distributions of comorbidities in a given population such as physical
health, cholesterol screening, diabetes, etc., are found to be
insightful in predicting outcomes \cite{Maleki}. Identifying such risk
factors and correlating them to the prediction accuracy provides
valuable candidates for future studies.

@@

A variety of mathematical and computational models have been used for
predicting COVID-19 outcomes \cite{Jewell}. It includes compartmental
models such as the SIR (Susceptible-Infected-Recovered) and SEIR
(Susceptible-Exposed-Infected-Recovered) which are relatively easy to
compute, and thus, commonly used in epidemiology. However, their
simplistic assumptions, such as homogeneous mixing within populations,
might make them less realistic for pandemics such as COVID-19. For
instance, such models often assume a closed population that does not
change. However, COVID-19 dynamics have prominently included massive
migration, deaths due to comorbidities temporarily conferred immunity,
and notably, major disparities in terms of socio-economic determinants
of health.

In terms of implementation, SIR and SEIR models are fitted with point
estimates, which may not adequately reflect the disease dynamics. For
instance, it is difficult to account for behavioral changes (e.g.,
human mobility) or systemic inadequacies (e.g., availability of
hospital beds) that may appear in specific subpopulations. Finally,
the compartmental model estimates do not allow for quantification of
uncertainty, which makes it harder to use them in policy-making. A
comparison of modeling approaches concluded that while SEIR model
performed better for a couple of states in the U.S., alternatives have
performed better for the other states \cite{Bertozzi}. AICov has
addressed most of the above modeling issues in its design, as per the
requirements of its forecasting objectives.

@@

\bibliographystyle{jds}
\bibliography{bib/data-covid19,bib/lstm,bib/cloudmesh}

>>>>

\usepackage{calc}
\usepackage{comment}
\usepackage{todonotes}
\setuptodonotes{inline}

@@

In this study, we used a Long Short-Term Memory (LSTM) algorithm for
our predictions \citep{www-keras-lstm,Hochreiter1997-dk}. An LSTM is a
recurrent neural network (RNN) \citep{Rumelhart1986-li} with feedback
connections allowing the use of data input sequences to predict data
output sequences. LSTMs have been applied to many application areas
from handwriting \citep{Graves2009-qb}, image, feature detection, and
time series prediction \citep{Schmidhuber2005-oy}. LSTM's have an
internal state that is used to prevent the vanishing gradient problem
\citep{Hochreiter1991-mp} during the training of RNNs.

@@

We often think of the laws of physics described by operators that
evolve the system given sufficient initial conditions and in this
language, we have shown how to represent Newton’s law operator by a
recurrent network \citep{Kadupitiya2020-zq}. We expect that the time
dependence of many complex systems: Covid pandemics, Southern
California earthquakes, traffic flow, security events can be described
by deep learning operators that both capture the dynamics and allow
predictions.

@@

\begin{requirement}{\bf Compute Resource Mapping.}
  For other uses of AICov, we need to integrate a flexible resource
  utilization framework. \Solution To address this requirement, we
  leverage from our earlier work and utilize through Cloudmesh and
  NIST NBDIF definitions cloud services into the architecture
  \citep{las-19-nist}. This includes containers but also infrastructure
  services in the cloud. Through these service integrations, we also
  provide access to GPUs.
\end{requirement}

@@

In summary, the AICov architecture allows the inclusion of new data
sources and models. An orchestrator enables modifying parameters and
selecting data sources and models used in the analysis. The framework
can be accessed via the interface layer that includes APIs, REST, and
Jupyter Notebooks. Compute interfaces are provided either by direct
access of GPUs from local resources or through the staging of
automatically generated REST services via our Cloudmesh REST Service
integrator
\citep{cloudmesh-openapi-install,cloudmesh-openapi-benchmark,cloudmesh-manual}. Additionally,
we can also access through abstractions cloud services that are
offered by the various cloud providers and are targeting our
application domain. Different models can be easily integrated into our
framework to verify our forecast or enhance our forecast based on the
users' models and data.

@@

While the U.S. has seen an overall case fatality rate of 5.14\%, some
U.S. locations have experienced a disproportionate number of deaths
compared to others. This disparity of deaths among different
communities could be attributed to a diversity of local risk factors
and socioeconomic determinants of health that now constitute a very
active area of investigation. Researchers are studying the effects of
different community-specific socioeconomic variables, underlying
health conditions and comorbidities, infrastructure, and systemic
responses.  Given the complexity of such underlying conditions that
can determine individual disease outcomes, where and under what
conditions people get infected by the virus, and then recover or die
take place not in isolation but rather in an interplay of diverse
factors that are characteristic to a local community, which need to be
synthesized. In other words, a basic requirement for effective
modeling of COVID-19 dynamics is that the information regarding a
particular community’s vulnerabilities and strengths pertaining to the
disease must be as comprehensive and integrative as possible.  The
regular reporting of accurate Coronavirus outbreaks data from local
health departments has been understandably difficult, especially in
the areas that are hardest hit. We use data on community-specific
health and behavioral risk factors aggregated at the county level from
the large-scale longitudinal CDC BRFSS surveys \citep{www-cdc-brfss},
and demographic and socioeconomic data from the U.S. Census Bureau
that are known at the local levels along with the cases and deaths
data compiled by NCHS \citep{www-cdc-nchs} and integrate them with
other data sources.

@@

Among the different sources of data on daily cases and deaths data
collected and aggregated at different spatial levels, especially for
the U.S., many are either freely available repositories such as GitHub
or accessible via specifically developed online resources such as the
New York Times~\citep{github-nytimes} and the Johns Hopkins Coronavirus
Resource Center~\citep{www-jh-covid19}.  For this study, we generated
two data sets based on the latter resource upon our checking for the
locations and times at which cumulative counts of COVID-19 cases and
deaths on any given day were no less than those on the previous
day. This led us to compile daily coronavirus cases and deaths data
for 345 U.S. counties from 32 states and the District of Columbia and
matched by five-digit FIPS code or county name to dynamic and static
variables from additional data sources.

The static data in this repository was collected from the U.S. Census
Bureau and the Centers for Disease Control and Prevention (CDC). The
different fields cover the domains of behavioral and health risk
factors, hospital capacity, and socioeconomic and demographic
conditions. Data on health outcomes, prevention measures, and
unhealthy behaviors were extracted from the CDC's 500 Cities Local
Data for Better Health program \citep{www-cdc-chronic-data} in the form
of crude or age-adjusted prevalence of conditions such as respiratory
disease, obesity or smoking. These values were aggregated from the
census tract to the county level using the first five digits of the
eleven-digit tract FIPS code. Because a single county may consist of
multiple individual cities, we include the list of all city labels
within each aggregate group to represent a greater metropolitan
area. 110 of such metropolitan areas that had more than 500 reported
cases of COVID-19 by April 15, 2020, were selected for this study.

@@

Demographic variables were gathered from the Census
QuickFacts~\citep{www-census} online resource using an automated web
scraping algorithm and cover relevant areas such as age, race, income,
and population density. Additional socioeconomic variables include the
Gini Index, which measures economic inequality and CDC Social
Vulnerability Index (SVI)~\citep{www-cdc-svi}. The SVI was created to
guide public health officials and disaster response efforts by
identifying the communities across the United States most likely to
need support during a crisis. Census tracts and corresponding counties
are ranked across 15 social factors, which are grouped into four
themes: socioeconomic status, household composition and disability,
minority status and language, and housing and transportation.

Lastly, the total number of general acute care, critical access, and
military hospitals within each county are included in the data. Such
variables included the number of relevant hospitals per county, and
the estimated number of beds (total known bed counts added to the
number of hospitals in the county with missing data times the average
number of beds per hospital in that state) per 1,000 people using the
American Hospital Directory~\citep{www-ahd}.

@@

To establish the feasibility of using deep learning, we have, in our
analysis, verified that through deep learning we can recreate the
predictions identified by a model-based approach using empirical
fits. In this comprehensive analysis, we obtain predictions through
deep learning while using data generated from epidemiological models
as discussed in \citep{marsland20-covid-paper}. The time series we used
was 100 days long, and a multi-layered Long Short-Term Memory (LSTM)
recurrent network was used. Our prediction approach differs by
learning not only from the demographics (fixed data for each city) and
time-dependent data but by integrating the population model for the
underlying prediction, as shown in Figure \ref{fig:magic-1}. Such model
predictive integration capability is important in any application with
multiple time scales. For example, this allows us to integrate
multi-scale time effects into the forecast, which could address the
combination of general forecasts and the next time steps of choice,
such as days, weeks, months, or longer. For this analysis, we used 37
of the 110 cities with reliable empirical fits (that we derived from
an epidemiological model and not deep learning
\citep{marsland20-covid-paper}) for the cases and deaths data up to
April 15, 2020. Our analysis identified a sophisticated single deep
learning time evolution operator that can describe these 37 separate
data sets, and smooth fitted data leads to very accurate deep learning
descriptions via LSTM as depicted in both Figure \ref{fig:magic-1}
that are smaller than 1\% from the model predictions.

@@

To answer these questions, we chose for our next analysis a date range
between 2020-02-01 and 2020-05-25. We start with a basic LSTM operator
with two layers of LSTM, and initial and final fully connected layers
\citep{Kadupitiya2020-zq}.  For LSTM, we have selected the activation
function based on Rectified Linear Units (RELU). For the last layer
sigmoid is chosen. A drop out rate of 0.2 is used, we use 16 input
nodes, and have a batch size of 110. The maximum epoch is chosen to be
200. The number of nodes internally is 32. The maximum total number of
Samples is 12540 when using 3 days as input length. We use a two-layer
LSTM, as shown in Table \ref{tab:model}. Within our system, we have
the ability to set a number of input days. For this analysis, we have
chosen three different input lengths, namely 3, 4, and 5 days.

@@

While this study focuses on if and which risk factors can improve the
predictions, we do not focus on why. Such analysis could provide
insights into causal relationships. For example, information such as
health insurance and the number of hospitals and beds are often
related to the quality of services that affect early detection and
treatment. Furthermore, population parameters such as density can
determine the rate of spread and the number of cases. Similarly,
distributions of comorbidities in a given population such as physical
health, cholesterol screening, diabetes, etc., are found to be
insightful in predicting outcomes \citep{Maleki}. Identifying such risk
factors and correlating them to the prediction accuracy provides
valuable candidates for future studies.

@@

A variety of mathematical and computational models have been used for
predicting COVID-19 outcomes \citep{Jewell}. It includes compartmental
models such as the SIR (Susceptible-Infected-Recovered) and SEIR
(Susceptible-Exposed-Infected-Recovered) which are relatively easy to
compute, and thus, commonly used in epidemiology. However, their
simplistic assumptions, such as homogeneous mixing within populations,
might make them less realistic for pandemics such as COVID-19. For
instance, such models often assume a closed population that does not
change. However, COVID-19 dynamics have prominently included massive
migration, deaths due to comorbidities temporarily conferred immunity,
and notably, major disparities in terms of socio-economic determinants
of health.

In terms of implementation, SIR and SEIR models are fitted with point
estimates, which may not adequately reflect the disease dynamics. For
instance, it is difficult to account for behavioral changes (e.g.,
human mobility) or systemic inadequacies (e.g., availability of
hospital beds) that may appear in specific subpopulations. Finally,
the compartmental model estimates do not allow for quantification of
uncertainty, which makes it harder to use them in policy-making. A
comparison of modeling approaches concluded that while SEIR model
performed better for a couple of states in the U.S., alternatives have
performed better for the other states \citep{Bertozzi}. AICov has
addressed most of the above modeling issues in its design, as per the
requirements of its forecasting objectives.

@@

\bibliographystyle{jdsurl}
\bibliography{bib/data-covid19,bib/lstm,bib/cloudmesh}
